---
title: "Repoducablt Research - Peer Assignment 1 "
author: "PK"
date: "July 17, 2015"
output: html_document
---
## Loading and preprocessing the data

```{r, echo = TRUE}
# Set the working directory and read the data
setwd("C:\\Users\\CECS-INT\\Desktop\\R_programs\\Reproducable Research")
mydata <- read.csv("activity.csv")
mydata3 <- data.matrix(mydata)
```


## What is mean total number of steps taken per day?
### Calculate the total number of steps taken per day
```{r, echo = TRUE}
# ignore NA's
mydata2 <- na.omit(mydata)
# Calculate the total number of steps taken per day
total_steps <- aggregate(x = mydata2$steps,
                         FUN = sum,
                         by = list(mydata2$date))
colnames(total_steps) = c("Date", "total_steps")

cat ("The total steps taken is")
total_steps
```

### Make a histogram of the total number of steps taken each day
``` {r, echo = TRUE}
hist(total_steps$total_steps, breaks = 30, xlab = "Total Steps", ylab = "Frequency", main = "Histogram of total steps taken each day" )
```


### Calculate and report the mean and median of the total number of steps taken per day
``` {r, echo = TRUE}
cat ("The mean of total number of steps taken per day is equal to ", mean(total_steps$total_steps))

cat ("The median of total number of steps taken per day is equal to ", median(total_steps$total_steps))
```

## What is the average daily activity pattern?

### Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

``` {r, echo = TRUE}
avg_steps <- aggregate(x = mydata2$steps,
                       FUN = mean,
                       by = list(mydata2$interval))
plot(avg_steps$Group.1, avg_steps$x, type = "l", xlab = "Interval", ylab = "Frequency", main = "Plot of the 5-minute interval vs average number of steps taken")
```

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
``` {r, echo = TRUE}
get.rows   <- (avg_steps$x == max(avg_steps$x))
max_avg_steps <- avg_steps[get.rows,]
cat ("The 5-minute interval that contains the maximum number of steps is", max_avg_steps$Group.1)
cat ("The number of steps taken during that duration is", max_avg_steps$x)
cat ("The time at that instant is", max_avg_steps$Group.1/100 )
```

## Imputing missing values

### Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

``` {r, echo = TRUE}
total_missing_values = nrow(mydata) - nrow(mydata2)
cat ("The total number of missing values is equal to ", total_missing_values)
```

### Devise a strategy for filling in all of the missing values in the dataset.


``` {r, echo = TRUE}
## I will use the mean of the 5 minute interval to fill the missing values 
mean_steps <- aggregate(x = mydata2$steps,
                         FUN = mean,
                         by = list(mydata2$interval))

```

### Create a new dataset that is equal to the original dataset but with the missing data filledin 

``` {r, echo = TRUE}
## The NA's are replaced by the mean of the interval they belong to 
for (i in 1:nrow(mydata3)){
  if(is.na(mydata[i,1,1])){
    get_r <- (mydata[i,,]$interval == mean_steps$Group.1)
     
     mdata <- mean_steps[get_r,]
     mydata3[i] <- mdata$x
  }
}
mydata3 <- data.frame(mydata3)
```

### Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day
``` {r, echo = TRUE}
total_steps_all <- aggregate(x = mydata3$steps,
                         FUN = sum,
                         by = list(mydata3$date))
colnames(total_steps_all) = c("Date", "total_steps")

hist(total_steps_all$total_steps, breaks = 30, xlab = "Total Steps", ylab = "Frequency", main = "Histogram of total steps taken each day")

cat ("The mean of total number of steps (replacing NA's) taken per day is equal to ", mean(total_steps_all$total_steps))

cat ("The median of total number of steps (replacing NA's) taken per day is equal to ", median(total_steps_all$total_steps))

```

###  Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
The mean reamined the same while the median increased by a negligible value. The impact of inputting the missing data on the estimates of the total daily number of steps is negligible at best 

## Are there differences in activity patterns between weekdays and weekends?
### Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
``` {r, echo = TRUE}
mydata$date <- as.Date(mydata$date)
df1 <- NA

  for (i in 1:nrow(mydata)){
    df1[i] <- weekdays(mydata$date[i])
  }
mydata3["Day"] <- df1

df2 <- subset(mydata3, Day %in% c("Saturday", "Sunday"))
df3 <- subset(mydata3, Day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))

avg_steps_weekday <- aggregate(x = df3$steps,
                         FUN = mean,
                         by = list(df3$interval))

avg_steps_weekend <- aggregate(x = df2$steps,
                               FUN = mean,
                               by = list(df2$interval))
avg_steps_weekday["Day_F"] <- "Weekday"
avg_steps_weekend["Day_F"] <- "Weekend"
avg_steps_allweek <- rbind(avg_steps_weekend,avg_steps_weekday)
```

###Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)
``` {r, echo = TRUE}
library(lattice)
xyplot(avg_steps_allweek$x ~ avg_steps_allweek$Group.1|avg_steps_allweek$Day_F, layout = c(1,2), type="l", xlab = "Interval", ylab = "Number of Steps" )

```
